\Large{\textbf{A biophysical, minimal model to investigate age-related changes in CA1 pyramidal cell electrical activity}}

### 2022-07-01 paper/article.tex
Update paper files

% long tables that run over 1pg
\usepackage{longtable}

@@

\affil[1]{\small{Departamento de F\'isica, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico}}
\affil[2]{\small{Laboratorio de Fisiolog\'ia de Sistemas, Departamento de Matem\'aticas, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico}}
\affil[3]{\small{Department of Psychology, Wilfrid Laurier University}}
\affil[4]{\small{McKnight Brain Institute, University of Arizona}}
\affil[*]{\small{Corresponding author: emckiernan@ciencias.unam.mx}}
\affil[**]{\small{Corresponding author: marcoh@ciencias.unam.mx}}

@@

For the dynamics of intracellular {\Ca}, we assume recovery toward a steady state $c_{\infty}$ at a rate $r_c$, with increments caused by the {\Ca} current $I_{Ca}$ \citep{avron1991minimal},
\begin{equation}
\partial_t{c} = r_{c}({c_{\infty} - c}) - \tilde{k}_{c} I_{CaL}(v,c).
\label{eq:dcdt}
\end{equation}
The term $\tilde{k}_c$ in equation~\eqref{eq:dcdt} is a conversion factor ($\mu$ M/pCoul) that accounts for the effect of {\Ca} flux across the membrane on the intracellular {\Ca} concentration.

@@

\paragraph{Change of variables to obtain numerical solutions.}
To simplify the numerics, we change variables $$u = v/v_T,$$ and adjust all voltages accordingly as $$u_l = v_l/v_T, \quad l \in \left\{NaT,CaL,K,NaK,m,n,w \right\}.$$ 
The new equation for the normalized voltage is $$\partial_t u = \frac{\partial_t v}{v_T}.$$
To simplify the notation and reduce the number of operations during the numerical integration, we also reparametrize the amplitudes as 
\begin{equation}A_l = \frac{2 a_l |\eta_l|}{v_T~Cm}, \label{eq:normA}
\end{equation} 
for $l \in \left\{NaT,CaL,K,NaK \right\}$, in units of 1/ms. 
Similarly,
the activation functions for the $NaT$, $CaL$, and $DK$ currents can also be rewritten respectively
as 
\begin{equation}
f_{\infty}(u) = \lrSet{1 + \exp\left[g_f \left(u_f-u \right) \right]}^{-1}, \quad{f \in  \left\{m,n,w \right\}}.
\end{equation}
The result is a new equation of the form
\begin{eqnarray}
\partial_t u &=& J_F -A_{NaT}(1-w)m_{\infty}(u) \varphi_{Na}(u) - A_{CaL}n_{\infty}(u) \varphi_{Ca}(u,c)- \notag \\
&& \left(A_{DK}w + A_{SK}H_{SK}(c)\right)\varphi_{K}(u)- A_{NaK}\varphi_{NaK}(u
).
\end{eqnarray}

@@

\begin{longtable}{p{0.09\textwidth} p{0.21\textwidth} p{0.1\textwidth} p{0.1\textwidth} p{0.35\textwidth}} 

@@

$a_{NaT}$ & amplitude of \newline transient {\Na} current & 1000-2300 & pA & set to produce currents of $\sim$2-7 nA, in the range recorded in CA1 PCs from rats \citep{ketelaars2001sodium} and guinea pigs \cite{sah1988sodium} \\
$a_{CaL}$ & amplitude of \newline L-type {\Ca} current & 25 or 50 & pA & set to produce currents of $\sim$2-3 nA or $\sim$5-6 nA as recorded in young and aged CA1 PCs, respectively \citep{campbell1996aging} \\
$a_{DK}$ & amplitude of delayed \newline rectifier {\K} current & 6000-8000 & pA & set to produce currents of $\sim$6-9 nA, in the range recorded from HEK cells expressing rat Kv2.1 and $I_K$ in hippocampal neurons \citep{mohapatra2009regulation} \\

@@

$a_{NaK}$ & amplitude of {\Na}/{\K} \newline ATPase current & 10-23 & pA & set to produce currents of $\sim$?-? pA, similar to but on high end of range recorded in hippocampal PCs \cite{richards2007differential} \\
$\eta_{x}$ & charge moved \newline across the membrane in a  single transport event  

@@

$\eta_{NaT}$ & charge transported \newline across the membrane by $NaT$ channels &-1& --& 1 net positive charge moving inward \cite{herrera2018thermodynamic} 

@@

$\eta_{CaL}$ & charge transported \newline across the membrane by Ca channels &-2& --& 2 net positive charges moving inward  \cite{herrera2018thermodynamic}

@@

@@

$r_{w}$ & rate of activation of \newline delayed rectifier {\K} \newline current & 1.0-1.8 & ms & 

@@

$s_{w}$ & symmetry of time \newline constant of delayed \newline rectifier {\K} current & 0.3 & - & based on fit; if higher (0.5-0.7) APs are the wrong shape and do not ride on sufficient plateau potential compared to recordings \\

@@

$c_{SK}$ & half-activation {\Ca} \newline concentration for SK \newline current & 740 & nM & based on recordings from oocytes expressing rat SK channel variant \citep{hirschberg1998gating,stocker20042+} \\
$g_{m}$ & activation slope of {\Na} \newline current & 5.0 & - & in range reported for {\Na} current in mouse \cite{carter2012transient} and rat \cite{costa1996kinetic} CA1 neurons \\
$g_{n}$ & activation slope of {\Ca} \newline current & 5.0 & - & \\
$g_{w}$ & activation slope of \newline delayed rectifier {\K} \newline current & 3.8& - & fit to data from rat brain delayed rectifier channels \cite{vandongen1990alteration} \\
$c_{\infty}$ & minimum intracellular {\Ca} concentration & 100 & nM & approximate resting intracellular {\Ca} \newline concentration in rat CA1 PCs \citep{oh2013altered,magee1996dihydropyridine,gant2006early} \\
$r_{c}$ & intracellular {\Ca} \newline removal rate constant &  1e$^{-3}$ \newline to \newline 5e$^{-3}$ & ms$^{-1}$ & adjusted to produce {\Ca} dynamics as recorded in rat CA1 PCs \citep{oh2013altered} \\
$k_{c}$ & conversion factor to \newline calculate effect of {\Ca} \newline current on intracellular {\Ca} concentration & 3e$^{-6}$ \newline to \newline 6e$^{-6}$ & nM/pCoul & adjusted to produce {\Ca} dynamics as recorded in rat CA1 PCs \citep{oh2013altered} \\

@@

% COMPARISON PARAMETER TABLE
\begin{table}[h!]
\centering
 \caption{Parameters used to produce different firing patterns in the young PC model. See equation~\eqref{eq:normA}.} 
\begin{footnotesize}
\rowcolors{2}{white}{gray!10}
\midsepremove
\begin{tabular}{p{9em} p{9em}  p{9em}  p{9em}} % spaces between the widths are important
\toprule
\textbf{parameter} & \textbf{adaptive \newline firing} & \textbf{conditional \newline bursting} & \textbf{spontaneous bursting} \\
\toprule
$a_{NaT}$ ($A_{NaT}$) & 1000 (2.99) & 1300 (3.89) & 2300 (6.88) \\
$a_{CaL}$ ($A_{CaL}$)& 25 (0.15) & 25 (0.15) & 25 (0.15) \\
$a_{DK}$ ($A_{KD}$)& 8000 (23.95) & 6000 (17.96) & 7000 (20.95) \\
$a_{SK}$ ($A_{SK}$)& 1400 (4.19) & 1600 (4.79) & 300 (0.90) \\
$a_{NaK}$ ($A_{NaK}$)& 10.0 (0.03) & 13 (0.04) & 23 (0.07) \\
$r_{w}$ & 1.0 & 1.8 & 1.1 \\
$r_{c}$ & 1e$^{-3}$ & 5e$^{-3}$ & 5e$^{-3}$ \\
$k_{c}$ & 3e$^{-6}$ & 6e$^{-6}$ & 6e$^{-6}$ \\
\bottomrule 
\end{tabular}
\vspace{0.2cm}
\begin{flushleft}
\footnotesize{Original amplitudes ($a_x$) are in pA. Reparametrized amplitudes ($A_x$) in parentheses are calculated by $2*\frac{a_x}{v_T C_m}$ for $x \in \left\{NaT,DK,SK,NaK\right\}$ and $4*\frac{a_{CaL}}{v_T C_m}$ for {\Ca} channels. For all calculations, $v_T C_m$ = 668.171 mv~pF.}
\end{flushleft}
\label{tab:regimes}
\end{footnotesize}
\end{table}
\midsepdefault

@@

\section*{Funding} 
This work was supported by grants from the Natural Sciences and Engineering Research Council of Canada, as well as the Ontario Mental Health Foundation, awarded to DFM. This work was also supported by DGAPA-UNAM-PAPIIT IA209817 awarded to ECM; and DGAPA-UNAM-PAPIIT IA208618 and IN228820, and DGAPA-UNAM-PAPIME PE114919, awarded to MAH-V.

>>>>

% long tables that run over 1pg
\usepackage{longtable}
\usepackage{array}

@@

\affil[1]{\small{Departamento de F\'isica, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico}}
\affil[2]{\small{Laboratorio de Din\'amica, Biof\'isica y Fisiolog\'ia de Sistemas, Departamento de Matem\'aticas, Facultad de Ciencias, Universidad Nacional Aut\'onoma de M\'exico}}
\affil[3]{\small{Department of Psychology, Wilfrid Laurier University}}
\affil[4]{\small{McKnight Brain Institute, University of Arizona}}
\affil[*]{\small{Corresponding author: emckiernan@ciencias.unam.mx}}
\affil[**]{\small{Corresponding author: marcoh@ciencias.unam.mx}}

@@

For the dynamics of intracellular {\Ca}, we assume recovery toward a steady state $c_{\infty}$ at a rate $r_c$, with increments caused by the {\Ca} current $I_{Ca}$ \citep{avron1991minimal},
\begin{equation}
\partial_t{c} = r_{c}({c_{\infty} - c}) - \tilde{k}_{c} I_{CaL}(v,c).
\label{eq:dcdt}
\end{equation}
The term $\tilde{k}_c$ in equation~\eqref{eq:dcdt} is a conversion factor ($\mu$M/pCoul) that accounts for the effect of {\Ca} flux across the membrane on the intracellular {\Ca} concentration.

@@

\paragraph{Change of variables to obtain numerical solutions.}
To simplify the numerics, we change variables $$u = v/v_T,$$ and adjust all voltages accordingly as $$u_l = v_l/v_T, \quad l \in \left\{NaT,CaL,K,NaK,m,n,w \right\}.$$ 
The new equation for the normalized voltage is $$\partial_t u = \frac{\partial_t v}{v_T}.$$
To simplify the notation and reduce the number of operations during the numerical integration, we also reparametrize the amplitudes as 
\begin{equation}A_l = \frac{2 a_l |\eta_l|}{v_T~C_m}, \label{eq:normA}
\end{equation} 
for $l \in \left\{NaT,CaL,K,NaK \right\}$, in units of 1/ms. 
Similarly,
the activation functions for the $NaT$, $CaL$, and $DK$ currents can also be rewritten respectively
as 
\begin{equation}
f_{\infty}(u) = \lrSet{1 + \exp\left[g_f \left(u_f-u \right) \right]}^{-1}, \quad{f \in  \left\{m,n,w \right\}}.
\end{equation}
The result is a new equation of the form
\begin{eqnarray}
\partial_t u &=& J_F -A_{NaT}(1-w)m_{\infty}(u) \varphi_{Na}(u) - A_{CaL}n_{\infty}(u) \varphi_{Ca}(u,c)- \notag \\
&& \left(A_{DK}w + A_{SK}H_{SK}(c)\right)\varphi_{K}(u)- A_{NaK}\varphi_{NaK}(u
).
\end{eqnarray}

@@

\begin{longtable}{p{0.09\textwidth} >{\raggedright\arraybackslash}p{0.21\textwidth} p{0.1\textwidth} p{0.1\textwidth} >{\raggedright\arraybackslash}p{0.37\textwidth}} 

@@

$a_{NaT}$ & amplitude of transient {\Na} current & 1000-2300 & pA & set to produce currents of $\sim$2-7 nA, in range recorded in CA1 PCs from rats \citep{ketelaars2001sodium} and guinea pigs \cite{sah1988sodium} \\
$a_{CaL}$ & amplitude of L-type {\Ca} current & 25 or 50 & pA & set to produce currents of $\sim$2-3 nA or $\sim$5-6 nA as recorded in young and aged CA1 PCs, respectively \citep{campbell1996aging} \\
$a_{DK}$ & amplitude of delayed rectifier {\K} current & 6000-8000 & pA & set to produce currents of $\sim$6-9 nA, in range recorded from HEK cells expressing rat Kv2.1 and $I_K$ in hippocampal neurons \citep{mohapatra2009regulation} \\

@@

$a_{NaK}$ & amplitude of {\Na}/{\K} ATPase current & 10-23 & pA & set to produce currents of $\sim$70-190 pA, similar to but on high end of range recorded in hippocampal PCs \cite{richards2007differential} \\
$\eta_{x}$ & charge moved across the membrane in a  single transport event  

@@

$\eta_{NaT}$ & charge transported across the membrane by $NaT$ channels &-1& --& 1 net positive charge moving inward \cite{herrera2018thermodynamic} 

@@

$\eta_{CaL}$ & charge transported across the membrane by Ca channels &-2& --& 2 net positive charges moving inward  \cite{herrera2018thermodynamic}

@@

$[Ca]_o$ & extracellular {\Ca} concentration (intracellular varies) & 1.5 & mM & in range reported for mammalian neural tissue \cite{egelman1999calcium} \\

@@

$r_{w}$ & rate of activation of delayed rectifier {\K} current & 1.0-1.8 & ms & 

@@

$s_{w}$ & symmetry of time constant of delayed rectifier {\K} current & 0.3 & - & based on fit; if higher (0.5-0.7) APs are the wrong shape and do not ride on sufficient plateau potential compared to recordings \\

@@

$c_{SK}$ & half-activation {\Ca} concentration for SK current & 7.4e$^{-4}$ & mM & based on recordings from oocytes expressing rat SK channel variant \citep{hirschberg1998gating} \\
$g_{m}$ & activation slope of {\Na} current & 5.0 & - & in range reported for {\Na} current in mouse \cite{carter2012transient} and rat \cite{costa1996kinetic} CA1 neurons \\
$g_{n}$ & activation slope of {\Ca} current & 5.0 & - & in range reported for Ca$_v$1.2 expressed in HEK cells \cite{shin2021sites}\\
$g_{w}$ & activation slope of delayed rectifier {\K} current & 3.8& - & fit to data from rat brain delayed rectifier channels \cite{vandongen1990alteration} \\
$c_{\infty}$ & minimum intracellular {\Ca} concentration & 1e$^{-4}$ & mM & equivalent to 100 nM, approximate resting intracellular {\Ca} concentration in rat CA1 PCs \citep{oh2013altered,magee1996dihydropyridine,gant2006early} \\
$r_{c}$ & intracellular {\Ca} removal rate constant &  1e$^{-3}$ \newline to \newline 5e$^{-3}$ & ms$^{-1}$ & adjusted to produce {\Ca} dynamics as recorded in rat CA1 PCs \citep{oh2013altered} \\
$k_{c}$ & conversion factor to calculate effect of {\Ca} current on intracellular {\Ca} concentration & 3e$^{-6}$ \newline to \newline 6e$^{-6}$ & mM & adjusted to produce {\Ca} dynamics as recorded in rat CA1 PCs \citep{oh2013altered} \\

@@

% COMPARISON PARAMETER TABLE
\begin{table}[h!]
\centering
 \caption{Parameters used to produce different firing patterns in the young PC model. See equation~\eqref{eq:normA}.} 
\begin{footnotesize}
\rowcolors{2}{white}{gray!10}
\midsepremove
\begin{tabular}{p{9em} p{9em}  p{9em}  p{9em}} % spaces between the widths are important
\toprule
\textbf{parameter} & \textbf{adaptive \newline firing} & \textbf{conditional \newline bursting} & \textbf{spontaneous bursting} \\
\toprule
$a_{NaT}$ ($A_{NaT}$) & 1000 (2.99) & 1300 (3.89) & 2300 (6.88) \\
$a_{CaL}$ ($A_{CaL}$)& 25 (0.15) & 25 (0.15) & 25 (0.15) \\
$a_{DK}$ ($A_{KD}$)& 8000 (23.95) & 6000 (17.96) & 7000 (20.95) \\
$a_{SK}$ ($A_{SK}$)& 1400 (4.19) & 1600 (4.79) & 300 (0.90) \\
$a_{NaK}$ ($A_{NaK}$)& 10.0 (0.03) & 13 (0.04) & 23 (0.07) \\
$r_{w}$ & 1.0 & 1.8 & 1.1 \\
$r_{c}$ & 1e$^{-3}$ & 5e$^{-3}$ & 5e$^{-3}$ \\
$k_{c}$ & 3e$^{-6}$ & 6e$^{-6}$ & 6e$^{-6}$ \\
\bottomrule 
\end{tabular}
\vspace{0.2cm}
\begin{flushleft}
\footnotesize{Original amplitudes ($a_x$) are in pA. Reparametrized amplitudes ($A_x$) in parentheses are calculated by $2*\frac{a_x}{v_T C_m}$ for $x \in \left\{NaT,DK,SK,NaK\right\}$ and $4*\frac{a_{CaL}}{v_T C_m}$ for {\Ca} channels. For all calculations, $v_T C_m$ = 668.171 mV~pF.}
\end{flushleft}
\label{tab:regimes}
\end{footnotesize}
\end{table}
\midsepdefault

@@

\section*{Funding} 
This work was supported by grants from the Natural Sciences and Engineering Research Council of Canada, as well as the Ontario Mental Health Foundation, awarded to DFM. This work was also supported by DGAPA-UNAM-PAPIIT IA209817 awarded to ECM; and by DGAPA-UNAM-PAPIIT IA208618, DGAPA-UNAM-PAPIIT IN228820, and DGAPA-UNAM-PAPIME PE114919 awarded to MAH-V.
